---
title: "CancerSignComp: Benchmarking Cancer Prognostic Gene Signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CancerSignComp: Benchmarking Cancer Prognostic Gene Signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Hundreds of prognostic gene signatures have been published for various cancer
types, yet direct comparisons between signatures — or against standard clinical
staging — are rarely performed in a standardized way. **CancerSignComp** fills
this gap by providing a unified framework for:

- Computing patient-level risk scores from gene expression matrices using
  weighted mean, GSVA, or ssGSEA methods
- Benchmarking multiple signatures against each other via concordance index
  (C-index) and log-rank test
- Comparing signatures to TNM/AJCC clinical staging using Net Reclassification
  Improvement (NRI) and Integrated Discrimination Improvement (IDI)
- Visualizing results with Kaplan-Meier curves, C-index bar plots, and subgroup
  forest plots

## Installation

```{r installation, eval = FALSE}
# Install from GitHub
devtools::install_github("bernahukkamli/CancerSignComp")
```

```{r setup}
library(CancerSignComp)
```

## Quick Start

The complete workflow requires four steps: load a signature, score patients,
compare signatures, and visualize results.

```{r quickstart}
set.seed(42)

# 1. Simulate a gene expression matrix (genes x patients)
mat <- matrix(
  rnorm(1000), nrow = 10,
  dimnames = list(
    c("TP53", "KRAS", "MYC", "CDKN2A", "SMAD4",
      "EGFR", "PTEN", "RB1", "BRCA2", "PIK3CA"),
    paste0("Patient_", 1:100)
  )
)

# 2. Simulate survival data
surv_time  <- rexp(100, rate = 0.05)
surv_event <- rbinom(100, 1, 0.7)

# 3. Define two signatures
sig1 <- load_signature(
  genes       = c("TP53", "KRAS", "MYC"),
  name        = "Signature_A",
  cancer_type = "PAAD"
)

sig2 <- load_signature(
  genes       = c("EGFR", "PTEN", "RB1"),
  name        = "Signature_B",
  cancer_type = "PAAD"
)

print(sig1)
```

```{r score}
# 4. Score patients using weighted mean (default)
scored1 <- score_signature(sig1, mat)
scored2 <- score_signature(sig2, mat)

print(scored1)
```

```{r compare}
# 5. Compare signatures
comp <- compare_signatures(list(sig1, sig2), mat, surv_time, surv_event)
print(comp)
```

## Visualization

### Kaplan-Meier Curves

```{r km, fig.cap = "Kaplan-Meier survival curves stratified by risk group for Signature A."}
p_km <- plot_kaplan_meier(scored1, surv_time, surv_event,
                          title = "Signature A — Risk Stratification")
print(p_km)
```

### C-index Comparison

```{r cindex, fig.cap = "C-index comparison across signatures. Dashed red line indicates random classifier (C-index = 0.5)."}
p_comp <- plot_comparison(comp,
                          title = "Signature Comparison: C-index")
print(p_comp)
```

### Subgroup Forest Plot

```{r forest, fig.cap = "Forest plot showing hazard ratios across patient subgroups."}
age_group <- sample(c("Young", "Old"), 100, replace = TRUE)

p_forest <- plot_forest(
  expr_matrix  = mat,
  surv_time    = surv_time,
  surv_event   = surv_event,
  signature    = sig1,
  subgroup_var = age_group,
  title        = "Signature A — Subgroup Analysis"
)
print(p_forest)
```

## Comparison to Clinical Staging

When TNM/AJCC stage information is available, `compare_to_staging()` tests
whether a signature provides prognostic value beyond clinical staging alone,
using NRI and IDI.

```{r staging}
# Simulate TNM stage (I–IV)
stage <- sample(1:4, 100, replace = TRUE, prob = c(0.15, 0.35, 0.35, 0.15))
staging_comp <- compare_to_staging(
  scored     = scored1,
  surv_time  = surv_time,
  surv_event = surv_event,
  stage      = stage
)

print(staging_comp)
```

## Scoring Methods

`score_signature()` supports three methods via the `method` argument:

| Method | Description | Best used when |
|--------|-------------|----------------|
| `"weighted_mean"` | Weighted average of gene expression | Signatures with known coefficients |
| `"gsva"` | Gene Set Variation Analysis | Pathway-level signatures |
| `"ssgsea"` | Single-sample GSEA | Large gene sets (≥ 15 genes) |

```{r methods, eval = FALSE}
# GSVA-based scoring
scored_gsva <- score_signature(sig1, mat, method = "gsva")

# ssGSEA-based scoring
scored_ssgsea <- score_signature(sig1, mat, method = "ssgsea")
```

## Saving Results

Results can be exported to CSV or Excel:

```{r save, eval = FALSE}
# Save to CSV
save_results(comp, "signature_comparison.csv")

# Save to Excel (requires openxlsx)
save_results(comp, "signature_comparison.xlsx")

# Save staging comparison
save_results(staging_comp, "staging_comparison.csv")
```


## Real-World Example: TCGA-LUAD Validation

This section demonstrates CancerSignComp using real TCGA-LUAD data
(n = 588 patients) to benchmark three published gene signatures against
TNM/AJCC clinical staging.

### Signatures Tested

| Signature | Genes | Cancer | Design Purpose |
|-----------|-------|--------|----------------|
| Zhou_2022 | COL1A1, GPX3, PLEK2 | LUAD | Prognostic |
| Zhang_2022 | SCAF11, NOD1, NLRP2, NLRP1, GPX4, CASP8, AIM2 | LUAD | Prognostic (pyroptosis) |
| Sarang_2025 | AGER, MGP, PECAM1, SLC2A1 | LUAD | Diagnostic |
```{r tcga_signatures, eval = FALSE}
# Define signatures
zhou_sig <- load_signature(
  genes  = c("COL1A1", "GPX3", "PLEK2"),
  name   = "Zhou_2022",
  source = "Front Oncol",
  cancer = "LUAD"
)

zhang_sig <- load_signature(
  genes  = c("SCAF11", "NOD1", "NLRP2", "NLRP1", "GPX4", "CASP8", "AIM2"),
  name   = "Zhang_2022",
  source = "Front Genet",
  cancer = "LUAD"
)

# Sarang_2025: diagnostic signature
# Downregulated genes get weight = -1, upregulated get weight = +1
sarang_sig <- load_signature(
  genes   = c("AGER", "MGP", "PECAM1", "SLC2A1"),
  name    = "Sarang_2025",
  source  = "Biomolecules",
  cancer  = "LUAD",
  weights = c(AGER = -1, MGP = -1, PECAM1 = -1, SLC2A1 = 1)
)
```

### Scoring and Benchmarking
```{r tcga_scoring, eval = FALSE}
# Score all three signatures
zhou_scored   <- score_signature(zhou_sig,   expr_matrix = expr_mat)
zhang_scored  <- score_signature(zhang_sig,  expr_matrix = expr_mat)
sarang_scored <- score_signature(sarang_sig, expr_matrix = expr_mat)

# Compare against each other
multi_comp <- compare_signatures(
  signatures  = list(zhou_scored, zhang_scored, sarang_scored),
  expr_matrix = expr_mat,
  surv_time   = os_time_months,
  surv_event  = os_status
)
print(multi_comp)
```

### Results

All three signatures were benchmarked against TNM/AJCC staging using
`compare_to_staging()`. Key findings:
```{r tcga_results, eval = FALSE}
sarang_comp <- compare_to_staging(
  scored     = sarang_scored,
  surv_time  = os_time_months,
  surv_event = os_status,
  stage      = stage_matched
)
print(sarang_comp)
```

The results highlight two important patterns:

**1. Prognostic signatures underperform TNM in well-staged cancers.**
Both Zhou_2022 (C-index = 0.552) and Zhang_2022 (C-index = 0.549)
failed to match TNM staging performance (C-index = 0.656) in LUAD,
despite being specifically designed as prognostic tools.

**2. Diagnostic signatures have no prognostic value.**
Sarang_2025 was originally designed to distinguish early-stage LUAD
from normal tissue. When repurposed as a prognostic tool, it showed
C-index = 0.511 (p = 0.145, ns) and HR = 0.95 (ns) <U+2014> statistically
indistinguishable from random classification. This exemplifies
"signature inflation": diagnostic signatures being uncritically
applied to prognostic contexts.

| Signature | C-index | Log-rank p | vs TNM (Delta) | Verdict |
|-----------|---------|------------|----------------|---------|
| Zhou_2022 | 0.552 | 0.005 | -0.096 | TNM > Sig |
| Zhang_2022 | 0.549 | 0.512 | -0.132 | TNM > Sig |
| Sarang_2025 | 0.511 | 0.145 | -0.145 | TNM > Sig |
| TNM staging | 0.656 | <0.001 | <U+2014> | Reference |

### Kaplan-Meier Visualization
```{r tcga_km, eval = FALSE}
# KM plot for Sarang_2025 (diagnostic signature <U+2014> no survival separation)
plot_kaplan_meier(
  scored     = sarang_scored,
  surv_time  = os_time_months[names(sarang_scored$risk_scores)],
  surv_event = os_status[names(sarang_scored$risk_scores)],
  title      = "Sarang_2025 - TCGA-LUAD (diagnostic signature)"
)
```

> **Interpretation:** The overlapping survival curves (p = 0.145) confirm
> that a signature optimized for tumor vs. normal tissue discrimination
> does not stratify patient prognosis. CancerSignComp makes this
> distinction explicit and quantifiable.
## Session Info

```{r sessioninfo}
sessionInfo()
```
